**Title:** Analyzing long-term water quality of lakes in Rhode Island and the Northeastern United States with an anomaly approach

### 2020-12-04 manuscript/ecosphere_resubmit_2/ecosphere_response_2.Rmd
edits on resubmit finished

General comments and thanks

@@

*Response:* Added description of regression line to caoptions and increased title text size.

@@

*Response:* Good catch.  The original language was confusing.  Caption now reads: "Panel A. Simulated, random data showing example of three hypothetical lakes; one lake was monitored throughout the record, one lake with low reported values monitored early in the record, and another lake with high reported values monitored late in the record.  Panel B. Analysis of simulated, random data with one lake monitored throughout the record, one lake with low reported values monitored early in the record, and another lake with high reported values monitored late in the record. Yearly average of the actual values is shown in red and yearly average of the site-specific anomalies is shown in blue."  Plus, have indicated in the methods (third paragaraph of Study Area and Data) that "lakes and reservoirs" are referred to as "sites" throughout the paper.

>>>>

Thank you for the review comments. The end product is significantly better.  We have itemized and responded to each of the comments below.  If there are any questions about what we have done, please do not hesitate to contact us.

@@

*Response:* Added description of regression line to captions and increased title text size.

@@

*Response:* Good catch.  The original language was confusing.  Caption now reads: "Panel A. Simulated, random data showing example of three hypothetical lakes; one lake was monitored throughout the record, one lake with low reported values monitored early in the record, and another lake with high reported values monitored late in the record.  Panel B. Analysis of simulated, random data with one lake monitored throughout the record, one lake with low reported values monitored early in the record, and another lake with high reported values monitored late in the record. Yearly average of the actual values is shown in red and yearly average of the site-specific anomalies is shown in blue."  Plus, we have indicated in the methods (third paragaraph of Study Area and Data) that "lakes and reservoirs" are referred to as "sites" throughout the paper.

### 2020-12-04 manuscript/ecosphere_resubmit_2/manuscript.Rmd
edits on resubmit finished

The study area for this analysis includes lakes and reservoirs in the state of Rhode Island where data were collected by the University of Rhode Island's Watershed Watch program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`) which allows for the examination of long-term trends in Rhode Island lakes.   The University of Rhode Island's Watershed Watch (URIWW) is a scientist-led community science program founded in the late 1980s that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.  Volunteer monitors are trained and then collect field data (e.g., sonde measurements or secchi depth) as well as whole water samples during the growing season (e.g., May through October) for later analysis in the lab.  The entire effort follows rigorous quality control/quality assurance protocols.  These types of community science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing; @Oliver_2017].

URIWW data contributed to the larger regional study by Oliver et al. [-@Oliver_2017], and, also allowed us to examine the long-term trends specifically in Rhode Island.  The URIWW program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring locations on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers. URIWW now provides more than 90% of Rhode Island's lake baseline data and is an integral part of the state's environmental data collection strategy. Data quality assurance and control is treated with paramount importance; volunteers are trained both in the classroom and the field, regular quality checks occur, and volunteers are provided with all the necessary equipment and supplies, along with scheduled collection dates. For freshwater lakes and reservoirs, weekly Secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and in deep lakes (greater than 5 meters) dissolved oxygen. Water samples are collected three times per season (May through October) to be analyzed for nutrients and bacteria.

For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.  In particular, we selected URIWW lakes and reservoirs (hereafter, referred to as sites) that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) had at least one sampling event per year in May through June, at least one sampling event per year in July through August, and at least one sampling event per year in September through October, 4) had at least one sampling event from 1993-2004 and 2005-2016, 5) and were sampled at a depth of 2 meters or less.  As not all sites have data for all selected years, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1993 to 2016 time frame.  The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs.  Of these sites, our filtered dataset had `r length(temp_10yr_sites)` sites measured for temperature, `r length(chla_10yr_sites)` sites measured for chlorophyll *a*, `r length(total_n_10yr_sites)` sites measured for TN, and `r length(total_p_10yr_sites)` sites measured for TP.  Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.  The N:P ratio was calculated by dividing the mass concentrations of total nitrogen and total phosphorus and then converting to a molar ratio by multiplying by 2.21 (e.g., atomic weight of P 30.974/atomic weight of N 14.007).

@@

Prior studies have modeled water quality trends across a larger region of the northeastern US within 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017]. We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al. [-@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017; @sorannolimno2019].  Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.  We used the same selection criteria on the LAGOSNE dataset as was applied to the URIWW data. This ensured that both datasets represented the same seasonal time frame.  Exact measurement depth for the LAGOSNE data were not available; however, the water quality data are reported to be either surface or epilimnion samples [@sorannolagosne2017]. At this time, the LAGOSNE data do not extend beyond 2013 thus sites needed to have a sampling event in 1993-2002 and in 2003-2013 to be included.

We use data, also from LAGOSNE, on landscape composition, maximum lake depth, and lake area [@sorannolagosne2017].  We use the 500 meter buffer landscape composition for three classes from the 2011 National Land Cover Dataset: Agriculture, Forest, and Developed.  The maximum lake depth is from various sources and lake area is calculated from the National Hydropgraphy Dataset waterbody polygons.  Additional details are available in Soranno et al. [-@sorannolagosne2017].

Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4050275> [@zenodo_citation].  All analyses were conducted with R version 4.0.3 and details on R package versions and operating system used for this analysis are included in a file, `sessioninfo.txt` at <https://github.com/USEPA/ri_wq_trends/blob/master/sessioninfo.txt> [@r2020; @lubridate; @broom; @stringr; @dplyr; @readr; @readxl; @tidyr; @here; @ggplot2; @hrbrthemes; @kendall; @sf; @usaboundaries; @usaboundariesdata; @knitr; @captioner; @pander; @cowplot; @tidycensus; @feddata; @nsink; @units; @xie2015dynamicdoc].  Values included in each of the figures have also been separately saved as a comma-separated values file, `yearly_average_anomaly.csv`, and may be retrieved via <https://github.com/USEPA/ri_wq_trends/raw/master/data/yearly_average_anomaly.csv>.

@@

There are many different methods for analyzing time series data for trends.  Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling sites is how to identify a trend while there is variation within a sampling location as well as variation introduced by differing start years for sampling among the many sites.  For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more degreaded water bodies.  Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].  The general approach we used calculates site-specific deviations from a long-term median over a pre-determined reference period.  This is slightly different than the typical use with temperature anomalies as those usually use the mean instead of the median.  Many of the variables we looked at are non-normal and often have outliers (e.g. algal blooms).  In this case the median is preferred. Using the median with site-specific deviations allows all sites to be shifted to a common baseline and the deviations, or anomalies, indicate change over the specified reference period.  We refer to this method as "site-specific anomalies".

@@

On average, the Rhode Island lakes and reservoirs show lower concentrations of nutrients with more nitrogen relative to phosphorus than in the larger LAGOSNE region (`r tab_num("ww_params", display = "cite")`, `r tab_num("lagos_params", display = "cite")`).  Chlorophyll concentrations also show lower average concentrations in Rhode Island (`r tab_num("ww_params", display = "cite")`, `r tab_num("lagos_params", display = "cite")`).  Furthermore, the distribution of chlorophyll based trophic state also shows a larger percentage of oligotrophic and mesotrophic lakes in Rhode Island and more eutrophic and hypereutrophic lakes in the larger LAGOSNE region [`r tab_num("ts_table", display = "cite", caption = "Percentage of lakes and reservoirs in different chlorophyll based trophic states.")`; @nurnberg1996trophic]. These results follow what would be expected given that, on average, LAGOSNE lakes are more dominated by agricultural lands than are Rhode Island lakes (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).

@@

Median annual trends for nutrients were not as clear. For total nitrogen, the slope of the line is positive with some, albeit weak, statistical support (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`) but the years greater than the median are distributed evenly throughout time.  Also, a slope of `r signif(ww_tn_rl$slope,2)` μg/L per year is lower than our threshold value of 3.5 μg/L suggesting little support for a meaningful ecological change over time (`r fig_num(name = "tn_trend", caption = "Twenty year trend for median TN anomaly.  Panel A. URI Watershed Watch yearly TN anomalies.  Panel B. LAGOSNE yearly TN anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.  Grey indicates yearly site-specific anomalies that were equal to the long-term medians.  Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.  Black line is the linear fit for year and median anomaly.", display = "cite")`A.). Also, 1998 only had three sites with available total nitrogen data and may not be representative.  This year also recorded the lowest median value.  Thus, 1998 may be skewing these results.  When this year is removed, the slope of the line is `r signif(ww_tn_rl_noearly$slope,2)` and the p-value is `r signif(ww_tn_rl_noearly$p.value, 2)`.  Thus, we interpret this as no trend in total nitrogen (`r tab_num(name = "trend_summ", display = "cite")`). Total phosphorus shows essentially no trend over time in the yearly anomalies with little statistical support (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over the median do not show any pattern (`r fig_num(name = "tp_trend", caption = "Twenty year trend for median TP anomaly.  Panel A. URI Watershed Watch yearly TP anomalies.  Panel B. LAGOSNE yearly TP anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.Grey indicates yearly site-specific anomalies that were equal to the long-term medians.  Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.  Black line is the linear fit for year and median anomaly.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`). Also a change of `r signif(ww_tp_rl$slope,2)` μg/L per year is much less than our defined threshold of 0.1 μg/L.  The TN:TP ratio has a small slope, very little statistical support, no pattern in the above and below years, and a slope magnitude less than 0.26 (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) suggesting little evidence for a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Twenty year trend for median TN:TP anomaly.  Panel A. URI Watershed Watch yearly TN:TP anomalies.  Panel B. LAGOSNE yearly TN:TP anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.Grey indicates yearly site-specific anomalies that were equal to the long-term medians.  Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.  Black line is the linear fit for year and median anomaly.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).

@@

In general, there was little evidence to suggest broad regional changes in chlorophyll *a* as it showed a very weak positive trend, slight statistical support and above average years spread evenly throughout the time period (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Total nitrogen showed a slight decreasing trend with weak statistical support and slope less than the 3.5 μg/L that would result in a trophic state change over a century, thus there is little support for a meaningful trend in TN at the regional scale (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Furthermore, The last two years for which LAGOSNE had data on total nitrogen were from a relatively small number of sites (`r tab_num("year_n_lagosne", display = "cite")`) and those values may not be representative.  Total phosphorus showed a very small increasing trend and there is statistical support for that trend.  The slope however suggests little support for an ecological meaningful change in total phosphorus (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`). Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`) as none of our criteria for a trend were met.  Although, there appears to be a non-linear trend that suggests increasing TN:TP early in the time period and decreasing TN:TP later, we feel we cannot effectively evaluate this though as total nitrogen measurements were rare in 2011 (`r tab_num("year_n_lagosne", display = "cite")`) and sites that had both nitrogen and phosphorus were even rarer.  This low median anomaly should be interpreted with this caveat in mind.   Taken together we feel our results largely match the findings of Oliver et al. [-@Oliver_2017] that there is stasis in lake nutrients and chlorophyll within the LAGOSNE region.

@@

Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent which were implemented in Rhode Island in 1995 [@riphosban1995; @litke1999review].  However, these nutrient reductions may not result in nutrient limitation and when when faced with other changes (e.g. temperature increases) increasing chlorophyll *a* may still occur.  Additionally, our analysis point to the fact that chlorophyll *a* and algal biomass is driven by processes operating at different scales.  For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].  Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as the percent of a catchment that is impervious surface and lake morphology can also impact temperature [@nelson2007stream]. In short, differences in regional and state level trends are driven by complex and multi-scale processes.

@@

Trends, and their ecological importance, could also vary depending on trophic state of a given waterbody.  We identified a minimum ecologically important change of 0.035 µg/l per year of chlorophyll.  This magnitude of change would result in a shift for any oligotrophic lake to a mesotrophic state; whereas, for any mesotrophic lake to become eutrophic over a century would require a change of 0.055 µg/l per year.  Examining these trends by trophic state provides the ability to look for ecologically important changes across the range of trophic state and not rely only on a single rate for all lakes.  Furthermore, Stoddard et al. [-@stoddard2016continental] report an 18.2% reduction in the number of oligotrophic lakes in the United States from 2007 to 2012.  A site-specific anomaly analysis by trophic state would identify trends, such as those seen in Stoddard et al. [-@stoddard2016continental], that might be occuring only on oligotrophic lakes.   A full analysis and interpretation of trends by trophic state is beyond the scope of this study.  However, we conducted a preliminary analysis of chlorophyll trends in Rhode Island for oligotrophic and hypereutrophic lakes which showed increasing chlorophyll trends for both trophic states but the yearly increase was much greater in the hypereutrophic lakes. This suggests that a full analysis of how lakes in different trophic states are changing over time is warranted.

@@

The site-specific anomalies are also robust to variations in sampling effort and in the timing of inclusion of given sampling locations.  For instance, if a site included only early in a time period had low values and a site included late in the time period had high values, then analyzing the measured values over this time period would show an increase whether or not one existed.  Using the site-specific anomalies rescales the values and allows the real trend to be seen.  We illustrate this with simulated, random data (e.g. no site-specific trend) (`r fig_num(name = "simulated", caption = "Panel A. Simulated, random data showing example of three hypothetical lakes; one lake was monitored throughout the record, one lake with low reported values monitored early in the record, and another lake with high reported values monitored late in the record.  Panel B. Analysis of simulated, random data with one lake monitored throughout the record, one lake with low reported values monitored early in the record, and another lake with high reported values monitored late in the record. Yearly average of the actual values is shown in red and yearly average of the site-specific anomalies is shown in blue.", display = "cite")`A.), and apply the site-specific analysis to those simulated data (`r fig_num(name = "simulated", display = "cite")`B.).  The average of the yearly measured values shows a trend (slope = `r signif(meas_lm$slope, 2)` and p-value = `r signif(meas_lm$p.value, 2)`), whereas the site-specific anomalies correctly showed no trend (slope = `r signif(anom_lm$slope, 2)` and p-value = `r signif(anom_lm$p.value, 2)`, `r fig_num(name = "simulated", display = "cite")`) as each site did not change over time.

>>>>

The study area for this analysis includes lakes and reservoirs in the state of Rhode Island where data were collected by the University of Rhode Island's Watershed Watch program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`) which allows for the examination of long-term trends in Rhode Island lakes.   The University of Rhode Island's Watershed Watch (URIWW) is a scientist-led community science program founded in the late 1980s that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.  Volunteer monitors are trained and then collect field data (e.g., sonde measurements or Secchi depth) as well as whole water samples during the growing season (e.g., May through October) for later analysis in the lab.  The entire effort follows rigorous quality control/quality assurance protocols.  These types of community science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing; @Oliver_2017].

URIWW data are included in the LAGOSNE and contributed to the larger regional study by Oliver et al. [-@Oliver_2017], which also allowed us to examine the long-term trends specifically in Rhode Island.  The URIWW program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring locations on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers. URIWW now provides more than 90% of Rhode Island's lake baseline data and is an integral part of the state's environmental data collection strategy. Data quality assurance and control is treated with paramount importance; volunteers are trained both in the classroom and the field, regular quality checks occur, and volunteers are provided with all the necessary equipment and supplies, along with scheduled collection dates. For freshwater lakes and reservoirs, weekly Secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and in deep lakes (greater than 5 meters) dissolved oxygen. Water samples are collected three times per season (May through October) to be analyzed for nutrients and bacteria.

For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.  In particular, we selected URIWW lakes and reservoirs (hereafter, referred to as sites) that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) had at least one sampling event per year in May through June, at least one sampling event per year in July through August, and at least one sampling event per year in September through October, 4) had at least one sampling event from 1993-2004 and 2005-2016, and 5) were sampled at a depth of 2 meters or less.  As not all sites have data for all selected years, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1993 to 2016 time frame.  The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs.  Of these sites, our filtered dataset had `r length(temp_10yr_sites)` sites measured for temperature, `r length(chla_10yr_sites)` sites measured for chlorophyll *a*, `r length(total_n_10yr_sites)` sites measured for TN, and `r length(total_p_10yr_sites)` sites measured for TP.  Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.  The N:P ratio was calculated by dividing the mass concentrations of total nitrogen and total phosphorus and then converting to a molar ratio by multiplying by 2.21 (e.g., atomic weight of P 30.974/atomic weight of N 14.007).

@@

Prior studies have modeled water quality trends across a larger region of the northeastern US within 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017]. We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al. [-@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017; @sorannolimno2019]. Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.  We used the same selection criteria on the LAGOSNE dataset as was applied to the URIWW data. This ensured that both datasets represented the same seasonal time frame.  Exact measurement depth for the LAGOSNE data were not available; however, the water quality data are reported to be either surface or epilimnion samples [@sorannolagosne2017]. At the time of our analysis, the LAGOSNE data do not extend beyond 2013 thus sites needed to have a sampling event in 1993-2002 and in 2003-2013 to be included.

We use data, also from LAGOSNE, on landscape composition, maximum lake depth, and lake area [@sorannolagosne2017].  We use the 500 meter buffer landscape composition for three classes from the 2011 National Land Cover Dataset: Agriculture, Forest, and Developed.  The maximum lake depth is from various sources and lake area is calculated from the National Hydrography Dataset waterbody polygons.  Additional details are available in Soranno et al. [-@sorannolagosne2017].

Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4050275> [@zenodo_citation].  All analyses were conducted with R version 4.0.3 and details on R package versions and operating system used for this analysis are included in a file, `sessioninfo.txt` at <https://github.com/USEPA/ri_wq_trends/blob/master/sessioninfo.txt> [@r2020; @lubridate; @broom; @stringr; @dplyr; @readr; @readxl; @tidyr; @here; @ggplot2; @hrbrthemes; @kendall; @sf; @usaboundaries; @usaboundariesdata; @knitr; @captioner; @pander; @cowplot; @tidycensus; @feddata; @nsink; @units; @xie2015dynamicdoc].  Values included in each of the figures have also been separately saved as a comma-separated values file, `yearly_average_anomaly.csv`, and may be retrieved via <https://github.com/USEPA/ri_wq_trends/blob/master/data/yearly_average_anomaly.csv>.

@@

There are many different methods for analyzing time series data for trends.  Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling sites is how to identify a trend while there is variation within a sampling location as well as variation introduced by differing start years for sampling among the many sites.  For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more degraded water bodies.  Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].  The general approach we used calculates site-specific deviations from a long-term median over a pre-determined reference period.  This is slightly different than the typical use with temperature anomalies as those usually use the mean instead of the median.  Many of the variables we looked at are non-normal and often have outliers (e.g. algal blooms).  In this case the median is preferred. Using the median with site-specific deviations allows all sites to be shifted to a common baseline and the deviations, or anomalies, indicate change over the specified reference period.  We refer to this method as "site-specific anomalies".

@@

On average, the Rhode Island lakes and reservoirs show lower concentrations of nutrients with more nitrogen relative to phosphorus than in the larger LAGOSNE region (`r tab_num("ww_params", display = "cite")`, `r tab_num("lagos_params", display = "cite")`).  Chlorophyll concentrations also show lower average concentrations in Rhode Island (`r tab_num("ww_params", display = "cite")`, `r tab_num("lagos_params", display = "cite")`).  Furthermore, the distribution of chlorophyll based trophic state also shows a larger percentage of oligotrophic and mesotrophic lakes in Rhode Island and more eutrophic and hypereutrophic lakes in the larger LAGOSNE region (`r tab_num("ts_table", display = "cite", caption = "Percentage of lakes and reservoirs in different chlorophyll based trophic states.")`; [@nurnberg1996trophic]). These results follow what would be expected given that, on average, LAGOSNE lakes are more dominated by agricultural lands than are Rhode Island lakes (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).

@@

Median annual trends for nutrients were not as clear. For total nitrogen, the slope of the line is positive with some, albeit weak, statistical support (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`) but the years greater than the median are distributed evenly throughout time.  Also, a slope of `r signif(ww_tn_rl$slope,2)` μg/L per year is lower than our threshold value of 3.5 μg/L suggesting little support for a meaningful ecological change over time (`r fig_num(name = "tn_trend", caption = "Twenty year trend for median TN anomaly.  Panel A. URI Watershed Watch yearly TN anomalies.  Panel B. LAGOSNE yearly TN anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.  Grey indicates yearly site-specific anomalies that were equal to the long-term medians.  Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.  Black line is the linear fit for year and median anomaly.", display = "cite")`A). Also, 1998 only had three sites with available total nitrogen data and may not be representative.  This year also recorded the lowest median value.  Thus, 1998 may be skewing these results.  When this year is removed, the slope of the line is `r signif(ww_tn_rl_noearly$slope,2)` and the p-value is `r signif(ww_tn_rl_noearly$p.value, 2)`.  Thus, we interpret this as no trend in total nitrogen (`r tab_num(name = "trend_summ", display = "cite")`). Total phosphorus shows essentially no trend over time in the yearly anomalies with little statistical support (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over the median do not show any pattern (`r fig_num(name = "tp_trend", caption = "Twenty year trend for median TP anomaly.  Panel A. URI Watershed Watch yearly TP anomalies.  Panel B. LAGOSNE yearly TP anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.Grey indicates yearly site-specific anomalies that were equal to the long-term medians.  Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.  Black line is the linear fit for year and median anomaly.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`). Also a change of `r signif(ww_tp_rl$slope,2)` μg/L per year is much less than our defined threshold of 0.1 μg/L.  The TN:TP ratio has a small slope, very little statistical support, no pattern in the above and below years, and a slope magnitude less than 0.26 (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) suggesting little evidence for a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Twenty year trend for median TN:TP anomaly.  Panel A. URI Watershed Watch yearly TN:TP anomalies.  Panel B. LAGOSNE yearly TN:TP anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.Grey indicates yearly site-specific anomalies that were equal to the long-term medians.  Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.  Black line is the linear fit for year and median anomaly.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).

@@

In general, there was little evidence to suggest broad regional changes in chlorophyll *a* as it showed a very weak positive trend, slight statistical support and above average years spread evenly throughout the time period (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Total nitrogen showed a slight decreasing trend with weak statistical support and slope less than the 3.5 μg/L that would result in a trophic state change over a century, thus there is little support for a meaningful trend in TN at the regional scale (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Furthermore, The last two years for which LAGOSNE had data on total nitrogen were from a relatively smaller number of sites (`r fig_num("year_n", display = "cite")`) and those values may not be representative.  Total phosphorus showed a very small increasing trend and there is statistical support for that trend.  The slope however suggests little support for an ecological meaningful change in total phosphorus (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`). Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`) as none of our criteria for a trend were met.  Although, there appears to be a non-linear trend that suggests increasing TN:TP early in the time period and decreasing TN:TP later, we feel we cannot effectively evaluate this though as total nitrogen measurements were rare in 2011 (`r fig_num("year_n", display = "cite")`) and sites that had both nitrogen and phosphorus were even rarer.  This low median anomaly should be interpreted with this caveat in mind.   Taken together we feel our results largely match the findings of Oliver et al. [-@Oliver_2017] that there is stasis in lake nutrients and chlorophyll within the LAGOSNE region.

@@

Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent which were implemented in Rhode Island in 1995 [@riphosban1995; @litke1999review].  However, these nutrient reductions may not result in nutrient limitation and when when faced with other changes (e.g. temperature increases) increasing chlorophyll *a* may still occur.  Additionally, our analysis point to the fact that chlorophyll *a* and algal biomass is driven by processes operating at different scales.  For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].  Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as the percent of impervious surface in a catchment and lake morphology can also impact temperature [@nelson2007stream]. In short, differences in regional and state level trends are driven by complex and multi-scale processes.

@@

Trends, and their ecological importance, could also vary depending on trophic state of a given waterbody.  We identified a minimum ecologically important change of 0.035 µg/l per year of chlorophyll.  This magnitude of change would result in a shift for any oligotrophic lake to a mesotrophic state; whereas, for any mesotrophic lake to become eutrophic over a century would require a change of 0.055 µg/l per year.  Examining these trends by trophic state provides the ability to look for ecologically important changes across the range of trophic state and not rely only on a single rate for all lakes.  Furthermore, Stoddard et al. [-@stoddard2016continental] report an 18.2% reduction in the number of oligotrophic lakes in the United States from 2007 to 2012.  A site-specific anomaly analysis by trophic state would identify trends, such as those seen in Stoddard et al. [-@stoddard2016continental], that might be occurring only on oligotrophic lakes.   A full analysis and interpretation of trends by trophic state is beyond the scope of this study.  However, we conducted a preliminary analysis of chlorophyll trends in Rhode Island for oligotrophic and hypereutrophic lakes which showed increasing chlorophyll trends for both trophic states but the yearly increase was much greater in the hypereutrophic lakes. This suggests that a full analysis of how lakes in different trophic states are changing over time is warranted.

@@

The site-specific anomalies are also robust to variations in sampling effort and in the timing of inclusion of given sampling locations.  For instance, if a site included only early in a time period had low values and a site included late in the time period had high values, then analyzing the measured values over this time period would show an increase whether or not one existed.  Using the site-specific anomalies rescales the values and allows the real trend to be seen.  We illustrate this with simulated, random data (e.g. no site-specific trend) (`r fig_num(name = "simulated", caption = "Panel A. Simulated, random data showing example of three hypothetical lakes; one lake was monitored throughout the record, one lake with low reported values monitored early in the record, and another lake with high reported values monitored late in the record.  Panel B. Analysis of simulated, random data with one lake monitored throughout the record, one lake with low reported values monitored early in the record, and another lake with high reported values monitored late in the record. Yearly average of the actual values is shown in red and yearly average of the site-specific anomalies is shown in blue.", display = "cite")`A), and apply the site-specific analysis to those simulated data (`r fig_num(name = "simulated", display = "cite")`B).  The average of the yearly measured values shows a trend (slope = `r signif(meas_lm$slope, 2)` and p-value = `r signif(meas_lm$p.value, 2)`), whereas the site-specific anomalies correctly shows no trend as each site did not change over time (slope = `r signif(anom_lm$slope, 2)` and p-value = `r signif(anom_lm$p.value, 2)`, `r fig_num(name = "simulated", display = "cite")`).
